---
jupyter:
  jupytext:
    notebook_metadata_filter: all,-language_info
    split_at_heading: true
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.16.6
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# An R tutorial on confidence intervals with plots

Here is a *population* of values.  In fact, in our case, it is a complete
collection of the birth weights for all children born in North Carolina in
2008.  See <https://github.com/odsti/datasets/tree/main/birth_weights> for more
on the source of the data.

```{r}
library(dplyr)
library(ggplot2)
library(cowplot)
```

```{r}
# Birth weights for every baby born in North Carolina.
bw_pop <- read.csv('data/nc_birth_weights.csv')
bw_pop_vals <- bw_pop[['birth_weight']]
pop_mean <- mean(bw_pop_vals)
```

But - we are very rarely in the situation where we know the values for the
entire population.   We are usually in the case where we have some random
sample of values from this population.

So, let's say we are in this typical situation, and we don't know anything
about the underlying population yet.  All we have is a single random sample of
50 values, that someone has taken for us.

```{r}
# A random sample of birthweights that someone has taken for us.
bw_sample <- read.csv('data/nc_birth_weights_sample.csv')
bw_sample_vals <- bw_sample[['birth_weight']]
sample_mean <- mean(bw_sample_vals)
```

```{r}
# The size of our sample(s).
n <- 50
```

```{r}
n_samples <- 4
row_2_start <- 2
row_3_start <- row_2_start + n_samples
layout_matrix <- rbind(c(1, numeric(n_samples - 1)),
                       row_2_start:(row_2_start + n_samples - 1),
                       row_3_start:(row_3_start + n_samples - 1))
layout(layout_matrix)
hist(bw_pop_vals)
abline(v = pop_mean, col = "blue", lwd = 3, lty = 3)
means <- numeric(n_samples)
breaks <- seq(0, 6, 0.5)
for (i in 1:n_samples) {
    sample <- sample(bw_pop_vals, size=50)
    sample_mean <- mean(sample)
    hist(sample, breaks=breaks, ylim=c(0, 25))
    abline(v = pop_mean, col = "blue", lwd = 3)
    abline(v = sample_mean, col = "red", lwd = 3)
    means[i] <- sample_mean
}
for (i in 1:n_samples) {
    hist(means)
}
```

```{r}

iris1 <- ggplot(iris, aes(x = Species, y = Sepal.Length)) +
  geom_boxplot() + theme_bw()

iris2 <- ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
  geom_density(alpha = 0.7) + theme_bw() +
  theme(legend.position = c(0.8, 0.8))

plot_grid(iris1, iris2, labels = "AUTO")

```

```{r}
tbw <- theme_set(theme_bw())
breaks <- seq(0.2, 5.5, by=0.2)

plot_bw <- function(df) {
    if (nrow(df) > 1000) {  # Population
        fill_col <- 'light grey'
        mean_col <- 'blue'
    } else {  # Sample
        fill_col <- 'dark grey'
        mean_col <- 'red'
    }
    mean_val <- mean(df[['birth_weight']])
    label <- paste('Mean:', sprintf('%.2f', mean_val))
    p <- (
        ggplot(df, aes(birth_weight))
        + geom_histogram(fill = fill_col, breaks = breaks)
        + geom_vline(aes(xintercept=mean_val, colour=label),
                     linewidth=1,
                     linetype='dashed')
        + labs(x = "Birth weight", colour="")  # Reset colour for later.
        + scale_colour_manual(values=mean_col)
        + theme(legend.position=c(0.25, 0.8),
                legend.text = element_text(size=20),
                aspect.ratio=0.7)
    )
}

p <- plot_bw(bw_pop) + labs(
    title = 'All NC birth weights',
    subtitle = '(Population)')
p
```

```{r}
ggsave('population.png', p)
```

Note that one can get axis limits like this:

```{r}
# https://stackoverflow.com/a/62819446
layer_scales(p)$y$get_limits()
```


```{r}
p <- plot_bw(bw_sample) + labs(
    title = 'Our sample')
ggsave('our_sample.png', p)
p
```

```{r}
set.seed(1939)
n <- 50
sample2 <- sample_n(bw_pop, n)
p <- plot_bw(sample2) + labs(
    title = 'A second sample')
ggsave('sample2.png', p)
sample3 <- sample_n(bw_pop, n)
p <- plot_bw(sample3) + labs(
    title = 'A third sample')
ggsave('sample3.png', p)
sample4 <- sample_n(bw_pop, n)
p <- plot_bw(sample4) + labs(
    title = 'A fourth sample')
ggsave('sample4.png', p)
```

Consider <https://wilkelab.org/cowplot/articles/shared_legends.html>
